Seismic parameter estimation and the Canadian crust

### 2013-07-18 paper/draftBP.tex
giant

%% -----------------------------------------------%%
\section{Data and methods}
  Analysis of bulk Canadian continental crust requires estimates of crustal properties such as seismic-wave velocity and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source for these estimates are data computed by processing raw seismograms into receiver functions and then inverting these RF's for parameter estimates. The other two datasets, the statistically compiled Crust 2.0 dataset and a compilation of pre-processed active source data, are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}

  The primary data utilized in this study are computed from teleseismic P-wave seismograms representing discrete seismic events. These events are comprised of seismic records representing more than 700 earthquake sources occurring between the years 2000 and 2012 at 343 broadband seismic stations across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Events are filtered by the epicentral distance from source to receiver with only the events within a 30 to 100 degree window being included. Seismic events are further filtered by hand picking those with reasonable signal to noise ratio and of sufficient impulsiveness that arrival times can be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

  The first stage of processing requires the transformation of teleseismic data into receiver functions. In its generic form this transformation involves deconvolving an approximation of the earthquake source from channels representing ground motion. The resulting waveform contains discrete pulses corresponding to the arrivals of S-wave energy scattered from subsurface discontinuities. Resolving these peaks is essential to the success of the following inversion, therefore it is advantageous to rotate the seismogram components to separate the P and S wave energy. This is accomplished by first rotating the N and E coordinates into radial and transverse dimensions and then performing a wave field decomposition on the radial and vertical channels [Bostock, 1998]. The direct arrival of the signal on the resulting P wave component is used as an approximation to the source function as P waves are simpler and have a response closer to that of a delta function. This windowed source coda is deconvolved from the S wave component computed from the wave field decomposition.

  An $L_2$ frequency domain deconvolution approach is used which has the advantages of computational efficiency as well as not requiring any assumptions about the noise in the data. This method performs a simultaneous deconvolution of $N$ seismograms sharing a similar slowness to compute a single impulse response or receiver function $r(t)$.
% Deconvolution equations
\begin{equation}
  r(t) = F^{-1} \left[ G(\omega) \right] = F{^-1}
 \left[ \frac {\sum_n^N S_n(\omega)P_n^*(\omega)} {\sum_n^N P_n(\omega)P_n^*(\omega) + \delta} \right ],
\end{equation}

@@

  A well tested and widely published method for extracting the ratio, $R=\frac{V_P}{V_S}$, where $V_P$ is P-wave velocity and $V_S$ is S-wave velocity and crustal thickness, or depth to Moho, $H$, is outlined by Zhu and Kanamori [2000], hereafter ZK. This method takes advantage of the differential arrival times between the S-wave reflected phases $Ps$, $PpPs$ $PsPs$, $PpSs$ and $PsSs$ and the direct P-wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ have an equal number of $P$ and $S$ phases, and the energy for these two phases arrive simultaneously and are indistinguishable. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

@@

  The travel time equations demand an apriori assumption on crustal P-wave velocity, $V_P$, and will trade-off to some degree with crustal thickness $H$. For this study, each station is assigned a $V_P$ value corrisponding to the Crust 2.0 value for the $2^o$ containing cell. In the specific cases where data is being compared to previously published results for quality control, $V_P$  values are chosen to match the values in the published study.

@@

  For a range of candidate models of $R$ and $H$ the RF's are stacked along trial moveout curves. During the stacking procedure each phase is assigned a weight to account for a general trend in the quality of the phases, with direct arrival usually carrying the best signal followed by $PpPs$ and $PpSs$. The weights chosen during the stacking are $w1 = 0.5$, $w2 = 0.3$ , $w3 = -0.2$ for the $Ps$, $PpPs$ and $PpSs$ phases respectively. A negative weight for the combined $PpSs$ and $PsPs$ phases is required as the polarity of the signal is reversed. Semblance weighting [Eaton, 2006] is employed to reduce the effect of spurious large amplitude noise in the data. The semblance function assigns a weight between zero (incoherent noise) and one (coherent signal). The stacking function is therefore defined as

@@

and $S_j$ is the semblance weight for the $j^{th}$ phase, time $t_j$ is calculated from the corrisponding travel time function for a given $H$, $R$ as a function of slowness $p_i$ and $N$ is the number of receiver functions. Multiplying by the semblance weighting sharpens the stacked image, $s(H,R)$, and exhibits better resolution when selecting between stacked models. The function $s(H,R)$ can be thought of as a transformation that maps the $Ps$, $PpPs$ and $PpSs+PsPs$ pulses of the receiver function, $r(t)$,  into $R$ and $H$ space as positive bands, via Eq. \ref{eq:stack}. Constructive interference where these bands intersect will produce a maximum in the stacking function (Fig \ref{fig:kinematicRvsH}). The model, $R$ and $H$, which produce this maximum provides the best estimate for the bulk crustal parameters for a given seismic station. The elongate shape about the point of intersection, with the long axis tilted mostly in the $R$ coordinate, conveys the inherent trade-off between $R$ and $H$ and the lower resolution exhibited by the seismic velocity ratio $R$.

@@

  Owing to increasing computational availability it is now reasonably efficient to extend the ZK approach and perform the stack along a range of all three seismic parameters, $V_P$, $R$ (therefore by definition $V_S$) and $H$. This has the advantage of not requiring an assumption on $V_P$ and should produce more accurate estimates, especially above crustal units with anomalous $V_P$. The inherent trade-off between $V_P$ and $R$ is limited, offering reasonable resolution between the two parameters. However, a significant trade-off between $V_P$ and $H$, which is a source of significant correlated error, requires that recovery of $V_P$ should be attempted only in cases where high frequency and high signal-to-noise ratio signals are present in the receiver function data. [[ Could go into more depth here, show the kinematic curves and poor trade-off between $V_P$ and $R$, Figure showing correlation between $V_P$ and $H$]]

  These trade-offs, along with other sources of error such as low signal-to-noise signals resulting from complex geology, are quantified by bootstrap resampling [Efron and Tibshirani, 1986]. Estimates are reprocessed using a random selection of receiver functions, with replacement, 1024 times. An estimate for the error is computed by taking the standard deviation from these trial runs.

\subsection{Additional data}
   Accompanying the processed estimates are data from controlled source experiments collected and compiled from GSC (Geological Survey of Canada) references by Walter Mooney (personal communication, 2012). This data provides $V_P$ and a few $V_S$ estimates for hundreds of experiment locations across Canada, many along traversals across geological provincial boundaries, faults and discontinuities. Some of these data are located within close enough proximity to seismic stations being utilized in this study that they may be used to compare with the $V_P$ and $V_S$ estimates resulting from the full parameter gridsearch.

  Whereas the active source data provide high resolution values at specific sites, the Crust 2.0 model, a global dataset widely cited in academic literature, provides complete coverage across Canada with an attendant trade-off in resolution. The data includes estimates of $V_P$ and $V_S$ for $2^o$ longitude and latitude cells as well as information regarding sedimentary thickness and crustal depth. While Crust 2.0 source data is a superset of the aforementioned active source data, the crustal model is also composed of statistically aggregated data from geographically distant but geologically similar regions. It therefore forms a unique dataset well suited for comparison with the crustal estimates calculated in this study.

@@

[[Comparison between $V_P$ estimates Mooney Active Source estimates]]

\subsection{Regional Bulk Crustal Parameters}
[[ Move from ZK + semblance to Full Grid Search + semblance ]]
  Data from X seismic stations across Canada are analyzed, using only those stations with less than 0.06 standard error. To compensate for the uneven geographic distribution of seismic stations, weights are applied to all crustal parameter estimates before averaging over a given region. Weights are calculated by first projecting station locations onto the 2D plane using the Albers equal-area conic projection. A Voronoi diagram is calculated for all projected station locations. The ratio between the Voronoi cell surface encompassing each station and the total area of the convex hull bounding region is computed and used as the weighting value.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{activeSourceComparison}
  \caption{Comparison of $V_P$ estimates from the MB algorithm with active source $V_P$ recordings. Stations selected if active source experiment locations within 1 degree of latitude / longitude.}
  \label{fig:activeComp}
\end{figure}

@@

\begin{bibliography}
Bostock, M. G. (1998), Mantle stratigraphy and the evolution of the Slave province, J. Geophys. Res., 103, 21,183-21,200.

@@

>>>>

%% -----------------------------------------------%%
\section{Data and methods}
Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source are teleseismic receiver functions (hereafter RFs). The remaining two datasets are Crust 2.0, a widely published statistical average of similar regions with a global two degree resolution and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 2.0 and the active source compilation are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic P-wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between the years 2000 and 2012 at and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Seismograms are included for analysis if they within a 30 to 100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with high enough signal to noise ratios to permit visual observation of the dominant P-wave energy and with sufficient impulsive first arrivals that allow arrival times to be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

The first stage of processing requires the transformation of teleseismic data into receiver functions. In general terms, this transformation involves deconvolving an approximation of the earthquake source from horizontal recordings of ground motion [Langston, 1979]. The resulting waveforms contain discrete pulses corresponding to the S-wave arrivals scattered from subsurface discontinuities including the base of the crust, or Moho. Confident identification of these peaks to the estimate of crustal parameters, and improved results can be obtained by transforming the seismogram channels into P and S contributions based on a 1-D model of propagation. This procedure is accomplished by first rotating the N and E coordinates into radial and transverse directions and then performing a wave field decomposition using the radial and vertical channels [Bostock, 1998]. The direct arrival of the signal on the resulting P wave component is used as an approximation to the source function as P-wave impulse response for teleseismic P approximates a delta function. This windowed source estimate is deconvolved from the S wave component computed from the wave field decomposition.

We employ $L_2$, multichannel, frequency domain approach to perform the deconvolution which has the advantage of computational efficiency and does not require a priori assumptions about the noise in the data. More specifically, we employ a simultaneous deconvolution of $N$ seismograms sharing a similar slowness to compute a single impulse response or receiver function $r(t)$.

% Deconvolution equations
\begin{equation}
  r(t) = F^{-1} \left[ G(\omega) \right] = F{^-1}
 \left[ \frac {\sum_n^N S_n(\omega)P_n^*(\omega)} {\sum_n^N P_n(\omega)P_n^*(\omega) + \delta} \right ],
\end{equation}

@@

A well tested and widely published method for extracting the $\frac{V_P}{V_S}$ ratio $R$ (where $V_P$ is P wave velocity and $V_S$ is S-wave velocity) and crustal thickness (or depth to Moho) $H$, is outlined by Zhu and Kanamori [2000], hereafter ZK. This method takes advantage of the differential arrival times between the S-wave reflected phases $Ps$, $PpPs$, $PsPs$, $PpSs$ and $PsSs$ and the direct P-wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ are kinematic analogues, such that the energy for these two phases arrive simultaneously for a 1-D model {\it COMMENT: note that the first phase is 2nd order whereas the 2nd is first
order in scattering strength, which is hwy PsPs is not always considered}. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by
ve

@@

The travel time equations, as employed by ZK, assume a value for crustal P-wave velocity, $V_P$, that
in practice will trade-off to some degree with crustal thickness $H$. In our implementation of the ZK approac, each station is assigned a $V_P$ value corrisponding to the Crust 2.0 value for the $2^\circ$ containing cell. In the specific cases where data are being compared to previously published results for quality control, $V_P$  values are chosen to match the values in the published study.

@@

RFs are stacked along trial moveout curves for a range of candidate models of $R$ and $H$ to generate a fitness function $s(H,R)$.  Within the stacking procedure, each phase is assigned a weight to account for a general trend in the quality of the phases, with direct conversion $Ps$ usually characterized by the highest quality signal followed by $PpPs$ and $PpSs$. We chose weights of $w1 = 0.5$, $w2 = 0.3$ , $w3 = -0.2$ for the $Ps$, $PpPs$ and $PpSs$ phases, respectively. A negative weight for the combined $PpSs$ and $PsPs$ phases is required as the polarity of the signal is reversed. Semblance weighting [Eaton, 2006] is employed to reduce the effect of spurious large amplitude noise in the data. The semblance function assigns a weight between zero (incoherent noise) and one (coherent signal). The stacking function is therefore defined as

@@

is the semblance weight for the $j^{th}$ phase, time $t_j$ is calculated from the corresponding travel time function for a given $H$, $R$ pair as a function of slowness $p_i$ and $N$ is the total number of receiver functions. Multiplying by the semblance weighting sharpens the stacked image, $s(H,R)$, and results in better resolution when discriminating between different models. The function $s(H,R)$ can be thought of as a transformation that maps the $Ps$, $PpPs$ and $PpSs+PsPs$ pulses of the receiver function, $r(t)$, into $R$ and $H$ space as positive bands, via Eq. \ref{eq:stack}. Constructive interference where these bands intersect will produce a maximum in the stacking function (Fig \ref{fig:kinematicRvsH}). The model, $R$ and $H$, which produce this maximum provides the best estimate for the bulk crustal parameters for a given seismic station. The elongate shape about the point of intersection, with the long axis tilted mostly in the $R$ coordinate, conveys the inherent trade-off between $R$ and $H$ and the lower resolution exhibited by the seismic velocity ratio $R$.

@@

In addition to employing the ZK approach, we also investigate the feasibility of separately estimating $V_P$ and $H$ through a full 3 parameter grid search involving these parameters and $R$. The trade-off between $V_P$ and $H$ (and other parameter combinations) and the effects of various errors, are quantified for both the full grid search and ZK methods by bootstrap resampling [Efron and Tibshirani, 1986]. Bootstrap standard deviations are produced using 1024 random selections of receiver functions, with replacement, per station.

\subsection{Additional data}

We supplement receiver function estimates of crustal properties with data from controlled source experiments collected and compiled from GSC (Geological Survey of Canada) references by Walter Mooney (personal communication, 2012). The active source data provide $V_P$ and a few $V_S$ estimates for hundreds of locations across Canada, many along transects that cross geological provincial boundaries, faults and discontinuities. Some of these data are located within reasonable ($<$100km) proximity to seismic stations used in this study and can be used to compare with the $V_P$ and $V_S$ estimates resulting from the RF full gridsearch.

The Crust 2.0 model [Bassin, 2000] employs many of these same active source data to provide regular, complete coverage across all of Canada. The model includes estimates of $V_P$ and $V_S$ information with depth at
$2^\circ$ resolution in latitude and longitude. In addition to employing local active source data, Crust 2.0 is also constrained by statistically aggregated data from geographically distant but geologically similar regions. It therefore affords a somewhat independent dataset suited for comparison with the crustal estimates calculated in this study.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{PCA}
  \caption{Trade-off between $H$ and $V_P$ captures by principal component analysis (PCA) of the full parameter search $V_P$ and $H$ estimates. The principal vector captures 89\% of the variance in the dataset and the correlation between $V_P$ and $H$ is 0.79. Independent random parameters will have zero correlation and principal components of roughly equal variance.}
  \label{fig:PCA}
\end{figure}

\subsection{Full Parameter Search Comparisons}
The primary limitation of the ZK approach is the requirement for an initial $V_P$ estimate. This can be resolved by computing the stacking function Eq. \ref{eq:stack}, as a function of all three seismic parameters, $s(H,R,V_P)$. Taking advantage of modern multi-core systems with reasonable quantities of data (fifty to a few hundred RFs) and a moderate search space ($n^3$ parameter candidates where $n \approx 150$) the computational time for this method can be measured in seconds.

Comparing the ZK approach with the full parameter search (Table \ref{table:ZKvsFG}) we see that the velocity ratio $R$ has a correlation between the two datasets of 0.76 while $H$ has a correlation of only 0.51. Since $t_{Ps}$ represents the differential travel time between the S and P wave, the dependence of $H$ on $V_P$ is weak. Error in $V_P$ will trade-off with $H$ (Figure \ref{fig:PCA} and can be removed by dividing out $V_P$ from $H$ in both datasets. The $H/V_P$ ratio between the two datasets shows a high correlation of 0.96 as expected.

If the error in $V_P$ is low the trade-off between $V_P$ and $H$ will be kept to a minimum. However, $V_P$ is difficult to resolve and even for clean stations the trade-off is quite large (Figure [slice through Vp space from 3D stack]). Taking the best three stations

\begin{table}
  \begin{tabular}{ l l }
    \hline
    Parameter & correlation coeff \\
    \hline
    $R$ (km) &  0.76 \\
    $H$      &  0.51 \\
    $H/V_P$  &  0.96 \\
    \hline
  \end{tabular}
  \caption{Correlation between parameters computed from ZK + semblance and full parameter search methods}
\label{table:ZKvsFG}

\subsection{Regional Bulk Crustal Parameters}
Data from X seismic stations across Canada are analyzed, using only those stations with less than 0.06 standard error. To compensate for the uneven geographic distribution of seismic stations, weights are applied to all crustal parameter estimates before averaging over a given region. Weights are calculated by first projecting station locations onto the 2D plane using the Albers equal-area conic projection. A Voronoi diagram is calculated for all projected station locations. The ratio between the Voronoi cell surface encompassing each station and the total area of the convex hull bounding region is computed and used as the weighting value.

@@

@@

\begin{bibliography}
Bassin, C., G. Laske, G. Masters (2000), The Current Limits of Resolution for Surface Wave Tomography in North America, EOS Trans AGU, 81.

Bostock, M. G. (1998), Mantle stratigraphy and the evolution of the Slave province, J. Geophys. Res., 103, 21,183-21,200.

@@

Langston, C. A. (1979), Structure under Mount Rainier, Washington, inferred from teleseismic body waves, J. Geophys. Res., 84(B9), 4749–4762, doi:10.1029/JB084iB09p04749.
